### 2016-05-30 text/Chapter02.tex
Edits and tweaks (Chapter 8 and 9 mostly)

We start (Section~\ref{sec:path-binding}) by discussing how coeffects arise from the most common
notion of context-dependence -- variable binding. Next, we look at coeffects as the dual of effect
systems (Section~\ref{sec:path-eff}) and we extend the duality to category theory, looking at
\emph{comonads} (Section~\ref{sec:path-sem}). Finally we consider logically inspired type systems
that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).

This chapter serves two purposes. Firstly, it provides a high-level overview of the  related work,
although technical details are often postponed until later. Secondly it recasts existing ideas in
a way that naturally leads to the coeffect systems developed later in the thesis. For this reason,
we are not always faithful to the referenced work -- sometimes we focus on aspects that the
authors consider unimportant or present the work differently than originally intended. The reason
is to fulfill the second goal of the chapter. When we do so, this is explicitly said in the text.

@@

\section{Through static and dynamic binding}
\label{sec:path-binding}

@@

In this section, we look at variable binding through the perspective of context-requirements.
We discuss ordinary variable binding and Haskell's implicit parameters \cite{app-implicit-parameters},
which provides an interesting point in the design space. We then discuss ambiguity that is inherent
in variable binding and, more generally, context-aware programming languages, and a way of
resolving it through \emph{type-directed semantics}.

@@

Variable access represents a form of context-dependence. For example, an expression $x+y$ can be
only evaluated if the environment provides values for variables $x$ and $y$. A variable
\emph{requirements} can be satisfied in two standard ways that are captured by \emph{dynamic} and
\emph{static} (or lexical) variable binding. Consider the following simple program:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The program can be evaluated in two ways, depending on the variable binding mechanism:
%
\begin{itemize}
\item \textsc{Static (lexical) binding.} In a language with static binding (such as ML or Java),
  the variable \ident{x} inside the body of the lambda function is statically bound to the
  declaration in the lexical scope -- that is, the variable on the second line -- and the
  expression evaluates to 10.

@@

In static and dynamic scoping, all variable requirements are always placed on one site. However,
those are not the only two options. It is conceivable that a language would use a mechanism that
splits variable requirements differently and combines aspects of dynamic and static binding. For
example, the language could use static binding by default, but resort to dynamic binding if a
variable is not available in the lexical scope. One such system is discussed in the next section.

Languages with static scoping generally check that all variable requirements are satisfied.
Attempt to access a variable that is not available in scope will give a compile error. In contrast,
languages with dynamic variable binding typically do not perform compile-time checks -- if a
program attempts to access a variable that is not available in the environment, a runtime error
occurs. This is possibly artifact of their implementation -- implementing static binding
\emph{without} checking would be cumbersome and implementing checking for dynamic binding is
requires a more sophisticated type system (Section~\ref{sec:path-effects-coeff}).

@@

The following two examples are variations on the one discussed in the Section~\ref{sec:path-binding-var}.
We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
implicit parameter is defined both at declaration-site and at the call site. On the right, the
implicit parameter is available only at the call site:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
delayed context demands and the expression evaluates to $10$. In the example on the right,
the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

In Haskell, the type system checks that bindings for all required implicit parameters are
available. The type of the function \ident{f} on the left is $\kvd{int}\rightarrow\kvd{int}$, while
the type of the \ident{f} function on the right is $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
in the environment when calling the function. It is worth noting that the syntax is similar to the
one used by type-class constraints. Those can be viewed as context demands too
(Section~\ref{sec:applications-flat-impl}).

\paragraph{Thesis perspective.}
The three different binding mechanisms discussed so far can be seen as different ways of splitting
context demands of a particular kind. Dynamic and static binding represent the opposite ends
of the design space and Haskell's implicit parameters are an interesting point inside the wider space.

In this thesis, we consider different notions of context, using implicit parameters as just one
of several motivating examples. However, implicit parameters are a valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
a function can be satisfied using the context available at declaration-site or using the context
available at the call site.

We also aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context demands are split between the declaration-site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one, different notions
of context might require different domain-specific choices and the general framework of
context-aware programming should make that possible.

@@

In many practical programming languages, the value and semantics of an expression depends on the
type derivation. Typically, the choice is hidden behind a mechanism that selects one preferred type
derivation.

This mechanism serves as an inspiration for our approach to resolving ambiguity inherent in
context-aware programs. This section shows a brief example using the F\# language \cite{app-fsharp},
before revisiting the implicit parameters example.

@@

The semantics of the function depends on the typing derivation. For example, for arrays, it is
compiled using the \ident{ldlen} intermediate language (IL) instruction, while or strings, it is
compiled using \ident{call} instruction (calling the property getter). In F\#, the compiler chooses
an appropriate typing derivation. For example:
%
\begin{equation*}
\begin{array}{l}
  \lbrack \str{hello};\str{world} \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
  \lbrack \ident{Array.empty};~\ident{Array.create}~\num{100}~\num{0}\; \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
\end{array}
\end{equation*}
%
The $|\hspace{-0.25em}>$ operator passes the value on the left to the function on the right. In
the first case, the compiler infers that the type of the input is a list of strings and so the
type of the lambda function becomes $\ident{string}\rightarrow\ident{int}$. In the second case,
the list contains two arrays (empty array and array containing one hundred 0 values) and so
the type of the lambda function is $\ident{int}\lbrack\,\rbrack\rightarrow\ident{int}$.
The important points about the example are:
%
\begin{itemize}
\item The semantics of the function $(\lambda\ident{x}\rightarrow\ident{x.Length})$
  depends on its type. For arrays, it is compiled using a special IL instruction, while for
  strings, it calls a property getter.

\item The compiler chooses an appropriate typing derivation. In the above case, this is done
  based on the context in which the expression appears, but other options are possible
  (in some cases, there is a \emph{default} resolution; in some cases, the compiler requires
  explicit typing annotation).
\end{itemize}
%
The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
possible typing derivations and its semantics varies depending on the type. If the lexical
scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
and it captures the value from the lexical scope. Otherwise, the type of the function is
$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
value from a hidden dictionary that is passed together with the input from the call site.

\paragraph{Thesis perspective.}
Just like the F\# function in the above example, certain expressions in context-aware languages
developed in this thesis have multiple valid typing derivations and their semantics depends on the
type. In F\#, the compiler determines a unique typing derivation based on other parts of the program
(or fails, if type is not uniquely determined). In our languages, we also determine a unique typing
derivation. However, rather than relying on type information from other parts of the program, we
explicitly define algorithm that chooses the preferred unique derivation.

This approach decouples two important aspects of context-aware programming and lets us study them
independently -- the semantics of context-aware programs and the domain-specific way of resolving
how context demands are satisfied. In our treatment of implicit parameters, we consider
multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
behaviour of Haskell).

@@

\section{Through type and effect systems}
\label{sec:path-eff}

@@

\subsection{Simple effect system.}
The structure of a simple effect system\footnote{Most work on effect systems uses $\sigma$ for
effect annotations. We use letters $\cclrd{r}, \cclrd{s}, \cclrd{t}$ and also distinguish effect
or coeffect annotations by colour.} is demonstrated in Figure~\ref{fig:path-eff}. The example
shows typing rules for a simply typed lambda calculus with an additional (effectful) operation
$l \leftarrow e$ that writes the value of $e$ to a mutable location $l$. The type of locations
($\ident{ref}_\rho~\tau$) is annotated with a \emph{memory region} $\rho$ of the location $l$.
The effects tracked by the type and effect system over-approximate the actual effects and memory
regions provide a convenient way to build such over-approximation. The effects are
represented as a set of effectful actions that an expression may perform and the effectful action
(\emph{write}) adds a primitive effect $\ident{write}(\rho)$.

@@

\section{Through language semantics}
\label{sec:path-sem}

@@

\paragraph{Language semantics.} When using monads to provide semantics of a language, we do not
need to extend the language in any way -- we assume that the language already contains the
effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
of the form $x\!:\!\tau_1 \vdash e : \tau_2$ is interpreted as a morphism $\tau_1 \rightarrow \mtyp{}{\tau_2}$,
meaning that any expression is interpreted as an effectful computation. The semantics of variable
access and the application of a primitive function $f$ is interpreted as follows:

\begin{equation*}
\begin{array}{rcl}
 \sem{x\!:\!\tau_1 \vdash x : \tau_1} &=& \ident{unit}_\mtyp{}{}\\
 \sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
\end{array}
\end{equation*}
%
Variable access is an effect-free computation, that returns the value of the variable, wrapped
using $\ident{unit}_\mtyp{}{}$. In the second rule, we assume that $e$ is an expression using
the variable $x$ and producing a value of type $\tau_2$ and that $f$ is a (primitive) function
$\tau_2 \rightarrow \mtyp{}{\tau_3}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
to a function $\mtyp{}{\tau_2} \rightarrow \mtyp{}{\tau_3}$ which is compatible with the
interpretation of the expression $e$.

@@

\paragraph{Meta-language interpretation.} ~When designing meta-language based on monads, we need to
extend the lambda calculus with additional type(s) and expressions that correspond to monadic
primitives:
%
\begin{align*}
\tau &:= \ident{num} \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
   e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
\end{align*}
%
The types consist of the primitive type, function type and a type constructor that
represents monadic computations. This means that the expressions in the language can create both
effect-free values, such as $\tau$ and computations $\mtyp{}{\tau}$. The additional expression
$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no actual effects) from a
value and $\kvd{let}_\mtyp{}{}$ is used to sequence effectful computations. In the semantics,
monads are not needed to interpret variable access and application, they are only used in the
semantics of additional (monadic) constructs:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \idf{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\tau_2}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\tau_3}} &=&
  \ident{bind}_\mtyp{}{}~\sem{e_2} \circ \sem{e_1}
\end{array}
\end{equation*}

@@

Now, consider a simple dataflow language with single-variable contexts, variables,
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$
has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
the language models dataflow and values are lists (of past values) is a matter of semantics,
which is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (extract the variable value); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).
For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an
expression that expects context with variable $x$ and returns a stream of values before the
previous one. The semantics of the term expresses this behaviour:
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed
earlier.

@@

\section{Through substructural and bunched logics}
\label{sec:path-logic}

@@

\section{Context oriented programming}

>>>>

We start (Section~\ref{sec:path-binding}) by discussing how coeffects arise from the most common
notion of context-dependence -- variable binding. Next, we look at coeffects as the dual of effect
systems (Section~\ref{sec:path-eff}) and we extend the duality to category theory, looking at
\emph{comonads} (Section~\ref{sec:path-sem}). We also consider type systems inspired by linear
and bunched logic, which are closely related to our structural coeffects (Section~\ref{sec:path-logic}).
Finally, we also consider practical motivations for context-aware programming (Section~\ref{sec:path-cop}).

This chapter serves two purposes. Firstly, it provides a high-level overview of the  related work,
although technical details are often postponed until later. Secondly it recasts existing ideas in
a way that naturally leads to the coeffect systems developed later in the thesis. For this reason,
we are not always faithful to the referenced work. We present the work through the coeffect view
and so we sometimes focus on aspects that authors consider unimportant or we present the work
differently than originally intended. When we do so, this is explicitly stated in the text.

@@

\section{Coeffects via static and dynamic binding}
\label{sec:path-binding}

@@

In this section, we look at variable binding through the perspective of context-requirements.
We discuss ordinary variable binding and Haskell's implicit parameters \cite{app-implicit-parameters},
which provide an interesting point in the design space. Implicit parameters give an example of
an ambiguity that arises more generally in context-aware programming, as well as one way of
resolving it through \emph{type-directed semantics}.

For a more general context-aware programming example, consider a program
running on a laptop that accesses an implicit parameter representing a printer. When printing, the
text may appear on my home printer (corresponding to static binding), or it may appear on the
nearest printer based on my physical location (corresponding to dynamic binding).

@@

Variable access represents a form of context-dependence. For example, an expression $x+y$ can be
only evaluated if the environment provides values for variables $x$ and $y$. A variable
\emph{requirements} can be satisfied in two standard ways that are characterized as \emph{dynamic}
and \emph{static} (or lexical) variable binding. Consider the following simple program:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The program can be evaluated in two ways, depending on the variable binding mechanism:
%
\begin{itemize}
\item \textsc{Static (lexical) binding.} In a language with static binding (such as ML or Java),
  the variable \ident{x} inside the body of the lambda function is statically bound to the
  declaration in the lexical scope -- that is, the variable on the second line -- and the
  expression evaluates to 10.

@@

In static and dynamic scoping, all variable requirements are always placed on one site. However,
those are not the only two options. It is conceivable that a language would use a mechanism that
splits variable requirements differently and combines aspects of dynamic and static binding. For
example, the language could use static binding by default, but resort to dynamic binding if a
variable is not available in the lexical scope. One such system is Haskell's implicit parameters
discussed in the next section.

Languages with static scoping resolve variable bindings at compile-time. In contrast,
languages with dynamic variable binding typically do not perform compile-time checks -- if a
program attempts to access a variable that is not available in the environment, a runtime error
occurs. This is possibly an artifact of their implementation -- implementing static binding
\emph{without} checking would be cumbersome and implementing static checking for dynamic binding is
requires a more sophisticated type system than traditional (Section~\ref{sec:path-effects-coeff}).

@@

The following two examples are variations on the one discussed in Section~\ref{sec:path-binding-var},
obtained by replacing a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
implicit parameter is declared both in the static scope and in the dynamic scope. On the right, the
implicit parameter is available only in the dynamic scope:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for Haskell's implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in static scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ and so the function \ident{f} has no
delayed context demands and thus the expression evaluates to $10$. On the right,
the context demand for \ident{?x} is \emph{delayed} and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

In Haskell, the type system checks that bindings for all required implicit parameters are
available and so no runtime errors can occur. The type of the function \ident{f} on the left is
$\kvd{int}\rightarrow\kvd{int}$, while the type of the \ident{f} function on the right is
$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
in the environment when calling the function. It is worth noting that the syntax is similar to the
one used by type-class constraints. Those can be viewed as context demands too
(Section~\ref{sec:applications-flat-impl}).

\paragraph{Thesis perspective.}
The three different binding mechanisms discussed so far can be seen as different ways of splitting
context demands of a particular kind into static and dynamic parts. Dynamic and static binding
represent the opposite ends of the design spectrum and Haskell's implicit parameters are an
interesting point inside the wider spectrum.

In this thesis, we consider various notions of context, using implicit parameters as one
of several motivating examples. Implicit parameters are a particularly valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
a function can be satisfied using the context available at declaration-site or using the context
available at the call site. Recall our earlier printer example -- a language that
provides access to resources in the context needs to provide enough flexibility in handling such
ambiguities, be it implicit parameter values available in scope or printers available in the
physical environment.

We aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context demands are split between the declaration-site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one for that particular
case, this thesis argues that other notions of context require different domain-specific choices
and the general framework of context-aware programming should make that possible.

@@

In many practical programming languages, the value and semantics of an expression depends on its
type derivation. In order to assign unique semantics to an expression, the choice is typically
hidden behind a mechanism that selects one preferred type derivation.

This mechanism serves as an inspiration for our approach to resolving ambiguity inherent in
context-aware programs. This section discusses a brief example using the F\# language \cite{app-fsharp},
before revisiting the implicit parameters example.

@@

The semantics of the function depends on the typing derivation. For example, for arrays, it is
compiled using the \ident{ldlen} intermediate language (IL) instruction, while or strings, it is
compiled using \ident{call} instruction (calling the property getter). In F\#, the compiler chooses
an appropriate typing derivation. For example:
%
\begin{equation*}
\begin{array}{l}
  \lbrack \str{hello};\str{world} \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
  \lbrack \ident{Array.empty};~\ident{Array.create}~\num{100}~\num{0}\; \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
\end{array}
\end{equation*}
%
The $|\hspace{-0.25em}>$ operator passes the value on the left to the function on the right. In
the first case, the compiler infers that the type of the input is a list of strings and so the
type of the lambda function becomes $\ident{string}\rightarrow\ident{int}$. In the second case,
the list contains two arrays (an empty array and an array containing one hundred zero values) and so
the type of the lambda function is $\ident{int}\lbrack\,\rbrack\rightarrow\ident{int}$.
The important points about the example are:
%
\begin{enumerate}
\item The semantics of the function $(\lambda\ident{x}\rightarrow\ident{x.Length})$
  depends on its type. For arrays, it is compiled using a special IL instruction, while for
  strings, it calls a property getter.

\item The compiler chooses an appropriate typing derivation. In the above case, this is done
  based on the context in which the expression appears, but other options are possible
  (in some cases, there is a \emph{default} resolution; often, the compiler requires an
  explicit typing annotation).

\item An expression without a type derivation does not have semantics. For example, given
  $\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})$, the F\# compiler fails to
  infer a type; the expression is not well-typed and does not have a semantics.\footnote{Alternatively,
  the compiler could choose default typing among multiple options. The F\# compiler does this
  for the $+$ operator, which can be used on \ident{float} and \ident{int} types, but the compiler
  chooses \ident{int} as the default.}
\end{enumerate}
%
The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
possible typing derivations and its semantics varies depending on its type. If the lexical
scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
and it captures the value from the lexical scope. Otherwise, the type of the function is
$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
value from a hidden dictionary that is passed together with the input from the call site.

\paragraph{Thesis perspective.}
Just like the F\# function in the above example, certain expressions in context-aware languages
developed in this thesis have multiple valid typing derivations and their semantics depends on the
type. In F\#, the compiler determines a unique typing derivation based on other parts of the program
(if type is not uniquely determined, it either chooses a default or fails). In our languages, we
also determine a unique typing derivation. However, rather than relying on type information from
other parts of the program, we explicitly define an algorithm that chooses the preferred unique
derivation (Section~Section~\ref{sec:flat-unique} and Section~\ref{sec:struct-unique}).

This approach decouples two important aspects of context-aware programming and lets us study them
independently -- the semantics of context-aware programs and the domain-specific way of resolving
ambiguities in how context demands are satisfied. In our treatment of implicit parameters, we consider
multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
behaviour of Haskell).

@@

\section{Coeffects via type and effect systems}
\label{sec:path-eff}

@@

\subsection{Simple effect system.}
The structure of a simple effect system\footnote{Most work on effect systems uses $\sigma$ or $F$ for
effect annotations. We use letters $\cclrd{r}, \cclrd{s}, \cclrd{t}$ and also distinguish effect
or coeffect annotations by colour.} is demonstrated in Figure~\ref{fig:path-eff}. The example
shows typing rules for a simply typed lambda calculus with an additional (effectful) operation
$l \leftarrow e$ that writes the value of $e$ to a mutable location $l$. The type of locations
($\ident{ref}_\rho~\tau$) is annotated with a \emph{memory region} $\rho$ of the location $l$.
The effects tracked by the type and effect system over-approximate the actual effects and memory
regions provide a convenient way to build such over-approximation. The effects are
represented as a set of effectful actions that an expression may perform and the effectful action
(\emph{write}) adds a primitive effect $\ident{write}(\rho)$.

@@

\section{Coeffects via language semantics}
\label{sec:path-sem}

@@

\paragraph{Language semantics.} When using monads to provide semantics of a language, we do not
need to extend the language in any way -- we assume that the language already contains the
effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
of the form $x\!:\!\tau_1 \vdash e : \tau_2$ is interpreted as a morphism $\tau_1 \rightarrow \mtyp{}{\tau_2}$,
meaning that any expression is interpreted as an effectful computation. The semantics of variable
access and the application of a primitive function $f$ is interpreted as follows:
%
\begin{equation*}
\begin{array}{rcl}
 \sem{x\!:\!\tau_1 \vdash x : \tau_1} &=& \ident{unit}_\mtyp{}{}\\
 \sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
\end{array}
\end{equation*}
%
Variable access is an effect-free computation, that returns the value of the variable, wrapped
using $\ident{unit}_\mtyp{}{}$. In the second rule, we assume that $e$ is an expression using
the variable $x$ and producing a value of type $\tau_2$ and that $f$ is a (primitive) function
$\tau_2 \rightarrow \mtyp{}{\tau_3}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
to a function $\mtyp{}{\tau_2} \rightarrow \mtyp{}{\tau_3}$ which is compatible with the
interpretation of the expression $e$.

@@

\paragraph{Meta-language interpretation.} ~When designing meta-language based on monads, we need to
extend the lambda calculus with additional type(s) and expressions that correspond to monadic
primitives:
%
\begin{align*}
\tau &:= \ident{num} \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
   e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
\end{align*}
%
The types consist of the primitive type, function type and a type constructor that
represents monadic computations. Thus the expressions in the language can create both
effect-free values, such as $\tau$ and computations $\mtyp{}{\tau}$. The additional expression
$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no effects) from a
value and $\kvd{let}_\mtyp{}{}$ is sequences effectful computations. In the semantics,
monads are not needed to interpret variable access and application, they are only used in the
semantics of additional (monadic) constructs:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \idf{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\tau_2}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\tau_3}} &=&
  \ident{bind}_\mtyp{}{}~\sem{e_2} \circ \sem{e_1}
\end{array}
\end{equation*}

@@

Now, consider a simple dataflow language with single-variable contexts, variables,
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$
has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
the language models dataflow and values are lists (of past values) is a matter of semantics,
which is defined as follows:

\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (extract the variable value); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).
For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an
expression that expects context with variable $x$ and returns a stream of values before the
previous one. The semantics of the term expresses this behaviour:
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed
earlier.

@@

\section{Coeffects via substructural and bunched logics}
\label{sec:path-logic}

@@

\section{Context oriented programming}
\label{sec:path-cop}
